Cartogram Mapping and its Application to Cancer Data Visualisation

### 2019-07-16 review_article.Rmd
section numbering

---
title: "Cartogram Mapping and its Application to Cancer Data Visualisation"
author: "Stephanie Kobakian, Jessie Roberts and Dianne Cook"
bibliography: reviewbib.bib
output:
  pdf_document: default
  word_document: default
  html_document: default
fontsize: 11pt
link-citations: yes
number_sections: yes
fig_caption: yes
biblio-style: vancouver
---

@@

## Abstract

@@

## Introduction
<!--
Current best practices: Choropleths
Current exploratory practices: Cartograms: cont, non cont, (tilegrams, dorlings)
Suggested uses of 
-->

## Disease mapping methods

### Current best practice map displays for cancer data

@@

<!-- what is a choropleth map -->
A choropleth map is used to show the spatial variability of measurements, with usage dating back to the 1800s. They are constructed by drawing administrative boundaries, and filling the polygon with colour to represent values of a measured variable. Early versions used symbols or patterns instead of colour. @CPISACA discusses the use of choropleth maps for visualising cancer data, and @DMAHP gives an overview of the development of the use of choropleth maps for displaying disease data. Figure 1 shows a choropleth map where each state of the United States of America is coloured by the average annual rate of new cases of lung and bronchus cancer from years 2012 to 2016. A choropleth map is a true map of the topology, constructed for visual inspection of spatial patterns across a familiar geographic form, that might show trends in disease occurrence, or even localised outbreaks.

@@

<!-- psychology -->
Utilising the familiar state shapes can make a map intuitive to read [@CIBMUK], and allow viewers to visually infer the spatial relationships in the data. 
The familiarity of the geography is a worthy consideration when presenting results of spatial analysis. Just as geographers are no longer the only creators of maps, @CPISACA suggests the audience of spatial health data analysis has extended beyond researchers to the public, policymakers and the media.
However, while they can be recognisable shapes, the different sizes of administrative areas can lead viewers to give more attention to the shades of the largest areas. Choropleths can inhibit visual inference when presenting human related statistics as the display may draw attention from the 'potentially more important results in the more populous communities' [@SE].

@@

### Public face

@@

#### Publishers {-}

@@

#### Reported measures {-}

@@

#### Geographical coverage {-}

@@

#### Statistical uncertainty

@@

A contiguous cartogram allows the map space to highlight the distribution of the variable. These displays keep spatial relationships of neighbours intact by preserving borders when adjusting sizes [@NAC].
From a computer graphics perspective, @ACA believe it is a problem of 'map deformation' to account for the value assigned to each area, they provide three methods for creating value-by-area cartograms.
A continuous area cartogram maintains connectivity of the map regions while areas are resized according to a statistic [@CBATCC].
Examples include Tobler’s Pseudo-Cartogram Method, Dorling’s Cellular Automaton Method [-@ACTUC], Radial Expansion Method of Selvin et al., Rubber Sheet Method of Dougenik et al., Gusein-Zade and Tikunov’s Line Integral Method, Constraint-Based Method (Kocmoud and House) [-@CBATCC].
- compare to their

@@

He answers this with his implementation of maps created with 'the simplest of all shapes'. While contiguous cartograms may be a 'more sophisticated' method, they produce 'very complex shapes'. Circular cartograms use the same shape for every region represented, and size them according to the population, or statistic represented.
To produce a compelling map, a gravity model is applied to avoid overlaps, and keep spatial relationships with neighbouring areas over many iterations. This implementation can work for up to 'one hundred thousand' areas.

@@

Used in displays of the UK by Howe in 1986 cited by @HEDP
<!--
## Temporal displays
-->

@@

### Centroid displays

@@

## A critique choropleth and cartograms

@@

## Animation

@@

## Acknowledgements

@@

## References

<!--
## Software
Commonly available tools to implement.
-->

<!--
### counts versus predictions

>>>>

---
title: "Cartogram Mapping and its Application to Cancer Data Visualisation"
author: "Stephanie Kobakian, Jessie Roberts and Dianne Cook"
bibliography: reviewbib.bib
output: 
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
  html_document: default
fontsize: 11pt
link-citations: true
fig_caption: yes
biblio-style: vancouver
---

@@

\section*{Abstract}

@@

# Introduction
<!--
Current best practices: Choropleths
Current exploratory practices: Cartograms: cont, non cont, (tilegrams, dorlings)
Suggested uses of 
-->

# Disease mapping methods

## Current best practice map displays for cancer data

@@

<!-- what is a choropleth map -->
A choropleth map is used to show the spatial variability of measurements, with usage dating back to the 1800s. They are constructed by drawing the given geographic or political boundaries, and filling the polygon with colour to represent values of a measured variable [@EI]. Early versions used symbols or patterns instead of colour. @CPISACA discusses the use of choropleth maps for visualising cancer data, and @DMAHP gives an overview of the development of the use of choropleth maps for displaying disease data. A choropleth map is a true map of the topology, constructed for visual inspection of spatial patterns across a familiar geographic form, that might show trends in disease occurrence, or even localised outbreaks. Figure 1 shows a choropleth map where each state of the United States of America is coloured by the average annual rate of new cases of lung and bronchus cancer from years 2012 to 2016.

@@

<!-- psychology -->
Utilising the familiar state shapes can make a map intuitive to read [@CIBMUK], and allow viewers to visually infer the spatial relationships in the data.
The familiarity of the geography is a worthy consideration when presenting results of spatial analysis. Just as geographers are no longer the only creators of maps, @CPISACA suggests the audience of spatial health data analysis has extended beyond researchers to the public, policymakers and the media.
However, while the areas are recognisable shapes, they are often politically driven boundaries with individual areas being of non-uniform size, different population densities and subject to changes over time. 
The different sizes of administrative areas can attract attention to the shades of the unpopulated, large physical areas [@EI]. Choropleths can inhibit visual inference when presenting human related statistics as the display may draw attention from the 'potentially more important results in the more populous communities' [@SE].

@@

## Public face

@@

### Publishers {-}

@@

### Reported measures {-}

@@

### Geographical coverage {-}

@@

### Statistical uncertainty

@@

A contiguous cartogram allows the map space to highlight the distribution of the variable. These cartograms maintain connectivity of the map regions while areas are resized according to a statistic, this often must occur at the expense of the shape [@CBATCC,@NAC].
From a computer graphics perspective, @ACA believe it is a problem of 'map deformation' to account for the value assigned to each area, they provide three methods for creating value-by-area cartograms.
Examples include Tobler’s Pseudo-Cartogram Method, Dorling’s Cellular Automaton Method [-@ACTUC], Radial Expansion Method of Selvin et al., Rubber Sheet Method of Dougenik et al., Gusein-Zade and Tikunov’s Line Integral Method, Constraint-Based Method (Kocmoud and House) [-@CBATCC].
- compare to their

@@

He answers this with his implementation of maps created with 'the simplest of all shapes'. While contiguous cartograms may be a 'more sophisticated' method, they produce 'very complex shapes'. Circular cartograms use the same shape for every region represented, and size them according to the statistic represented or the population for a base map.
To produce a compelling map, a gravity model is applied to avoid overlaps, and keep spatial relationships with neighbouring areas over many iterations. This implementation can work for up to 'one hundred thousand' areas.

@@

Used in displays of the UK by Howe in 1986 cited by @HEDP
<!--
# Temporal displays
-->

@@

## Centroid displays

@@

# A critique of mapping methods

@@

# Animation

@@

# Acknowledgements

@@

# References

<!--
# Software
Commonly available tools to implement.
-->

<!--
## counts versus predictions
